Harmony: A Verifiable Systems Programming Language

### 2022-01-16 paper/paper.tex

<{:end:}>
\end{tcolorbox}
}
\\
(a) & (b)
\end{tabular}
\end{center}
\caption{(a) Lock specification and (b) a spinlock implementation based on
test-and-set}
\label{fig:synch}
\end{figure*}

@@

\subsection{Basics}

For example, \autoref{fig:ms} shows a reproduction of the well-known
Michael\&Scott two-lock concurrent queue~\cite{MS96} along with a
Harmony version of the same algorithm.
Harmony uses the `<{?}>' instead of the `<{&}>' symbol to take the address
of a memory location, but inherits the same <{q->field}> notation
from C to access the given field in the dictionary pointed to by
pointer <{q}>.  <{q->field}> is shorthand for <{(!q)["field"]}> in
Harmony, where `<{!}>' is the Harmony operator to dereference a pointer.
Harmony also complains, correctly, about a data race when
there is a concurrent \texttt{enqueue} and \texttt{dequeue} operation
on an empty queue---the implementer should ensure that accesses to the
\texttt{next} field are sequentially consistent.

@@

Statements in Harmony, including groups of statements, can be tagged with
the \texttt{atomically} keyword to make them atomic.  What this means
is that the statement is evaluated atomically by the Harmony virtual
machine without interleaving of instructions of other threads.  It is
a \emph{stronger} condition than holding a lock---a lock still allows
concurrent execution of multiple threads, but it prevents multiple
threads from executing in the same critical section.
The <{atomically}> keyword is intended for <{specification}> more than
for <{implementation}> of concurrent algorithms.

For example, \autoref{fig:synch}(a) shows the specification for locks
from the Harmony <{synch}> module that was used in \autoref{fig:ms},
featuring the Harmony <{when}> statement.
\autoref{fig:synch}(b) shows an implementation of the <{acquire}>
method based on a test-and-set primitive, which in turn is
\emph{specified}.
Imprecisely, methods that use the <{atomically}> keyword are specifications.
Whether something is a specification or an implementation is in the
eye of the beholder.
Important, however, is the notion of modularity that is implied here.
Harmony can model check the implementation of \autoref{fig:ms} using
the specification of a lock in \autoref{fig:synch}(a).
Separately, Harmony can check the correctness of the implementation in
\autoref{fig:synch}(b) against the specification in
\autoref{fig:synch}(a).

@@

This process is automated in Harmony.  The approach is to
run the model checker twice.  The first time, the test program is
run against the specification of the concurrent queue.  Harmony
outputs a DFA such as shown in \autoref{fig:helloworld}.  The second
time, the same test program is run against the implementation of the
concurrent queue, but with the inclusion of the DFA.  The model
checker ends up verifying the Cartesian product of the implementation
and the DFA and ensures that every behavior of the implementation
program (in terms of its print statements) is also a behavior of
the specification program.  In case the implementation has a behavior
that is not allowed by the specificatino, the model checker can
provide the shortest counter-example.

\autoref{fig:qcmp} shows some measurements taken from running
Harmony against the test program in \autoref{fig:queue}(b).
The Harmony compiler allows both constants and modules to be
specified on the command line, so the test program can easily be
run with different versions of the <{queue}> module and different
values for <{NOPS}>, the number of threads spawned to
perform a queue operation.  The table has a row for three different
configurations and a column for four different values for <{NOPS}>.
It is clear that running the test program against the specification
involves significantly fewer states, and therefore less time, than
running the test program against an implementation, particularly
if the behavior of that implementation must be checked as well.

In practice, <{NOPS = 4}> is probably sufficient to check the
correctness of a concurrent queue implementation.  Note, however,
that if there are more than two operations to check, the model
checking time will also increase.  In such cases, it usually suffices
to run all possible pairs of read-only and read-write operations.

@@

Harmony can be used to model a distributed system in which the
only shared variable is a set of messages representing the network.
A message is sent by atomically adding a message to the network:
<{def send(m): atomically network |= {m}}> (`<{|}>' is the union
operator in Python).
To receive a message, a program can use the Harmony <{when}>
statement to wait for a message.  Typically, however, the program
does not want to receive just any message, but a message that satisfies
certain conditions such as: \emph{wait for the next message that is addressed
to me and has a sequence number larger than 3}.
For this, Harmony supports the <{exists}> clause and allows
<{when}> clauses to be nested:
<{atomically when exists m in network when (m.dst == self) and (m.seq > 3): ...}>
With these constructs, Harmony can be effectively used as a
model checker for distributed algorithms.

@@

\paragraph{Stop/Go.}  Harmony supports continuations as a first-class
construct.  To enable this, \texttt{Stop} saves the current context
(with its program counter incremented) in a specified shared location
(such as the tail of a scheduling queue) and removes the context
from the context bag and active set.  If the context was atomic
(which is usually the case), the active set is replaced with the
domain of the context bag.  Even though its atomic counter is not
zero, the stopped context is not considered atomic at this time,
but will resume being atomic once it is scheduled to run.

@@

\paragraph{Print.}  \texttt{Print} pops a value of the stack and increments
the program counter.  In case a behavior automaton has been specified, the
value is applied to the DFA state and checked for validity.
To reduce state explosion, the HVM does not keep track of the value itself
in the state (like keeping the value in a shared list as part of the state).
Instead, Harmony keeps printed values on the edges between the states
in the Kripke structure.

@@

@@

\begin{figure*}
\begin{center}
\includegraphics[width=.9\textwidth]{diners.png}
\end{center}
\caption{The HTML output of running the Dining Philosophers program}
\label{fig:dinersHTML}
\end{figure*}

When Charm finds a problematic state or transition, it outputs
a detailed trace that is called a \emph{counter-example}.
Harmony always outputs the shortest such trace.
For ease of debugging, traces are ordered first by the number of
context switches they make, and then by the number of states on
the path.
However, in order to save memory, Charm does not actually keep
the required information during the construction of the Kripke
structure.  In particular, its edges only contain the context
and the choice if any.  Thus, in order to reconstruct the
counter-example, Charm runs the HVM one more time, this time
re-evaluating the edges along the path and outputting the
detailed information about each transition.

@@

If Charm determines a problem, it outputs the shortest counter-example that
it could find.  However, this counter-example is at the level of the HVM,
not the Harmony program that the programmer wrote.  It can be hard to
dig through a list of HVM instructions.
Harmony includes tools to quickly focus on the root cause of the
problem without having to look at the HVM instructions themselves.
In particular, in case of a problem, Harmony outputs a compact
but detailed and interactive overview of the problem in HTML format that
can be accessed from any web browser.

@@

While Harmony has proven itself to be effective and even fun to use
for a large variety of synchronization problems and distributed
algorithms, it is not a panacea.  At the core of Harmony is a model
checker and---even with all its optimizations---it can still suffer
from combinatorial state explosion in programs that have too many
threads or too many non-deterministic choices.  Nonetheless, because
of its systematic state exploration, many problems can often be
uncovered in small scenarios with few threads and few choices.
Harmony's support for anonymous threads is useful for scaling when
there are many similar threads that do not need to know their own
identifier.  Also, Harmony's support for modular model checking
allows larger systems to be built from smaller ones.

The Harmony language, at least currently, is not usable for production
code.  One reason for this is the <{atomically}> clause.  On a multi-core
processor it would require an instruction to stop all other cores.
Harmony was not designed to be a language for production code.  Because
it is close to languages such as Python and C, it tends to be fairly
easy to hand-translate code developed in Harmony to a desired target
programming language.  However, such hand-translation could easily
introduce a bug that did not exist in the Harmony code.

@@

PlusCal~\cite{Lamport09} is an imperative programming language that,
after translation to TLA+~\cite{Lamport02} (a formal verification
language), can be run through the TLC model checker.  Harmony also
supports compilation to TLA+.  While similar in spirit to Harmony,
Harmony first compiles to a virtual machine that is close to CPUs
like x86, ARM, and RISC-V, using a stack and common control flow
techniques.  The resulting HVM has a TLA+ specification.  PlusCal,
instead, \emph{trans\-piles} to TLA+.  Informally, the gap between
PlusCal programs and their TLA+ representations is much smaller
than the gap between Harmony programs and their TLA+ representations,
making PlusCal more amenable to formal verification.  On the other
hand, working effectively with PlusCal does require significant
knowledge of TLA+, increasing the learning curve.

>>>>

<{:end:}>
\end{tcolorbox}
}
\\
(a) & (b)
\end{tabular}
\end{center}
\caption{(a) Lock specification and (b) a spinlock implementation based on
a specification of test-and-set}
\label{fig:synch}
\end{figure*}

For example, \autoref{fig:ms} shows a reproduction of the well-known
Michael\&Scott two-lock concurrent queue~\cite{MS96} along with a
Harmony version of the same algorithm.
Harmony uses the `<{?}>' instead of the `<{&}>' symbol to take the address
of a memory location, but inherits the same <{q->field}> notation
from C to access the given field in the dictionary pointed to by
pointer <{q}>.  <{q->field}> is shorthand for <{(!q)["field"]}> in
Harmony, where `<{!}>' is the Harmony operator to dereference a pointer.
Harmony complains, correctly, about a data race when
there is a concurrent \texttt{enqueue} and \texttt{dequeue} operation
on an empty queue---the implementer should ensure that accesses to the
\texttt{next} field are sequentially consistent.

@@

@@

Statements in Harmony, including groups of statements, can be tagged with
the \texttt{atomically} keyword to make them atomic.  What this means
is that the statement is evaluated atomically by the Harmony virtual
machine without interleaving of instructions from other threads or interrupt
handlers.  It is
a \emph{stronger} condition than holding a lock---a lock still allows
concurrent execution of multiple threads, but it prevents multiple
threads from executing in the same critical section.
The <{atomically}> keyword is intended for <{specification}> more than
for <{implementation}> of concurrent algorithms.

For example, \autoref{fig:synch}(a) shows the specification for locks
from the Harmony <{synch}> module that was used in \autoref{fig:ms},
featuring the Harmony <{when}> statement.
\autoref{fig:synch}(b) shows an implementation of the <{acquire}>
method based on a test-and-set primitive, which in turn is
\emph{specified}.
Imprecisely, methods that use the <{atomically}> keyword are specifications.
Important is the notion of modularity that is implied here.
Harmony can model check the implementation of \autoref{fig:ms} using
the specification of a lock in \autoref{fig:synch}(a).
Separately, Harmony can check the correctness of the implementation in
\autoref{fig:synch}(b) against the specification in
\autoref{fig:synch}(a).

@@

This process is automated in Harmony.  The approach is to
run the model checker twice.  The first time, the test program is
run against the specification of the concurrent queue.  Harmony
outputs a DFA such as shown in \autoref{fig:helloworld}.  The second
time, the same test program is run against the implementation of the
concurrent queue, but with the inclusion of the DFA.  The model
checker ends up verifying the Cartesian product of the implementation
and the DFA and ensures that every behavior of the implementation
program (in terms of its print statements) is also a behavior of
the specification program.  In case the implementation has a behavior
that is not allowed by the specification, the model checker can
provide the shortest counter-example.

\autoref{fig:qcmp} shows measurements taken from running
Harmony against the test program in \autoref{fig:queue}(b).
The Harmony compiler allows both constants and modules to be
specified on the command line, so the test program can easily be
run with different versions of the <{queue}> module and different
values for <{NOPS}>, the number of threads spawned to
perform a queue operation.  The table has a row for three different
configurations and a column for four different values for <{NOPS}>.
It is clear that running the test program against the specification
involves significantly fewer states, and therefore less time, than
running the test program against an implementation, particularly
if the behavior of that implementation must be checked as well.

In practice, <{NOPS = 4}> is probably sufficient to check the
correctness of a concurrent queue implementation.  Note, however,
that if there are more than two operations to check, the model
checking time will also increase.  In such cases, it usually suffices
to run all pairs of read-only and read-write operations.

@@

Harmony can be used to model a distributed system in which the
only shared variable is a set of messages representing the network.
From a model checking perspective, it is convenient that message and
timer handling can be performed atomically, as the only interaction
between threads is through the network variable.
A message is sent by atomically adding a message to the network:
<{def send(m): atomically network |= {m}}> (`<{|}>' is the union
operator in Python).
To receive a message, a program can use the Harmony <{when}>
statement to wait for a message.  Typically, however, the program
does not want to receive just any message, but a message that satisfies
certain conditions such as: \emph{wait for the next message that is addressed
to me and has a sequence number larger than 3}.
For this, Harmony supports the <{exists}> clause and allows
<{when}> clauses to be nested:
<{atomically when exists m in network when (m.dst == self) and (m.seq > 3): ...}>
With these constructs, Harmony can be effectively used as a
model checker for distributed algorithms.

@@

\paragraph{Stop/Go.} \texttt{Stop} saves the current context
(with its program counter incremented) in a specified shared location
(such as the tail of a scheduling queue) and removes the context
from the context bag and active set.  If the context was atomic
(which is usually the case), the active set is replaced with the
domain of the context bag.  Even though its atomic counter is not
zero, the stopped context is not considered atomic at this time,
but will resume being atomic once it is scheduled to run.

@@

\paragraph{Print.}  \texttt{Print} pops a value of the stack and increments
the program counter.  In case a behavior automaton has been specified, the
value is applied to the DFA state and checked for validity.
To reduce state explosion, the HVM does not keep track of the value itself
in the state (like keeping the value in a shared list as part of the state).
Instead, Harmony keeps printed values on the edges between the states.

@@

When Charm finds a problematic state or transition, it outputs
a detailed trace that is called a \emph{counter-example}.
Harmony always outputs the shortest such trace.
For ease of debugging, traces are ordered first by the number of
context switches they make, and then by the number of states on
the path.
However, in order to save memory, Charm does not actually keep
the required information during the construction of the Kripke
structure.  In particular, its edges only contain the context,
whether an interrupt occurred, and the choice if any.
Thus, in order to reconstruct the
counter-example, Charm runs the HVM one more time, this time
re-evaluating the edges along the path and outputting the
detailed information about each transition.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{diners.png}
\end{center}
\caption{The HTML output of running the Dining Philosophers program}
\label{fig:dinersHTML}
\end{figure*}

@@

If Charm determines a problem, it outputs the shortest counter-example that
it could find.  However, this counter-example is at the level of the HVM,
not the Harmony program that the programmer wrote.  Without help it can be
difficult to dig through a list of HVM instructions.
Harmony includes tools to quickly focus on the root cause of the
problem without having to look at the HVM instructions themselves.
In particular, in case of a problem, Harmony outputs a compact
but detailed and interactive overview of the problem in HTML format that
can be accessed from any web browser.

@@

While Harmony has proven itself to be effective and even fun to use
for a large variety of synchronization problems and distributed
algorithms, it is not a panacea.  At the core of Harmony is a model
checker and---even with all its optimizations---it can still suffer
from combinatorial state explosion in programs that have too many
threads or too many non-deterministic choices.  Nonetheless, because
of its systematic state exploration, many problems can often be
uncovered in small scenarios with few threads and few choices.
Harmony's support for anonymous threads is useful for scaling when
there are many similar threads that do not need to know their own
identifier.  Also, Harmony's support for modular model checking
allows implementing larger systems using the specifications of smaller ones.

Currently there is no compiler that generates production code from
a Harmony program.
Because it is close to languages such as Python and C, it tends to be fairly
easy to hand-translate code developed in Harmony to a desired target
programming language.  However, such hand-translation could easily
introduce a bug that did not exist in the Harmony code.
A Harmony compiler would have to be able to recognize constructs such as
the test-and-set specification in \autoref{fig:synch}(b) and generate
a hardware test-and-set instruction.

@@

PlusCal~\cite{Lamport09} is an imperative programming language that,
after translation to TLA+~\cite{Lamport02} (a formal verification
language), can be run through the TLC model checker.  Harmony also
supports compilation to TLA+.  While similar in spirit to Harmony,
Harmony first compiles to a virtual machine that is close to CPUs
like x86, ARM, and RISC-V, using a stack and common control flow
techniques.  The resulting HVM has a TLA+ specification.  PlusCal,
instead, \emph{trans\-piles} to TLA+.  Informally, the gap between
PlusCal programs and their TLA+ representations is much smaller
than the gap between Harmony programs and their TLA+ representations,
making PlusCal programs more amenable to formal verification.  On the other
hand, working effectively with PlusCal does require significant
knowledge of TLA+, increasing the learning curve.
